---
title: "Education and BP, ANOVA analyses"
author: "Adriana Lääti et al."
date: "`r Sys.Date()`"
output: html_document
---
 

```{r setup, include=FALSE} 

# Load packages
library(pacman)
library(tableone)
library(officer)
library(flextable)
library(knitr)
p_load(here, dplyr)

# Use package "here" to load data sets in a cross-platform (PC, Mac etc.) way:
#
# First, tell where you are right now - i.e. location of the file you are running right now

here::i_am("article2/README.R")

# After that, you can check where the project root is by running "here()"
here()

#  Load the datasets using {here} package
article2tuva_file <- here("data", "tuva_edu.rds")
article2utuva_file <- here("data", "utuva_edu.rds")
article2tuva_utuva_file <- here("data", "tuva_utuva_edu.rds")


tuva <- readRDS(article2tuva_file)
utuva <- readRDS(article2utuva_file)

tuva_utuva <- readRDS(article2tuva_utuva_file)

# Clean up variables
rm(article2tuva_file, article2utuva_file, article2tuva_utuva_file)
 


# Load hypertension prevalence and BP medication data from Article 1 analyses
ht_file_tuva <- here("data", "article1tuva_bp_meds_final.rds")
ht_file_utuva <- here("data", "article1utuva_bp_meds_final.rds")

ht_tuva <- readRDS(ht_file_tuva)
ht_utuva <- readRDS(ht_file_utuva)

# Clean up variables
# rm(article2tuva_file, article2utuva_file)
rm(article2file, ht_file_tuva, ht_file_utuva)

# Only retain the wanted hypertension/medication variables
ht_tuva <- ht_tuva %>% select(ID1991, hypertension_tuva1, antihypertensive_med)
ht_tuva <- ht_tuva %>% rename(hypertension = hypertension_tuva1)

ht_utuva <- ht_utuva %>% select(ID1991, antihypertensive_med, hypertension_utuva1)
ht_utuva <- ht_utuva %>% rename(hypertension = hypertension_utuva1)

# Merge the hypertension prevalence and BP medication use data from Article 1
tuva <- left_join(tuva, ht_tuva, by = "ID1991")


utuva <- left_join(utuva, ht_utuva, by = "ID1991")



# Clean up variables
# rm(article2tuva_file, article2utuva_file)
rm(ht_tuva, ht_utuva)
  
# Set up necessary general settings
knitr::opts_chunk$set(echo = FALSE)





```
 
 
<br>
<br> 
 
 

## Display N number for TUVA data

<br>

```{r n_tuva}

tuva_utuva %>% filter(cohort == "1920-born TUVA cohort") %>% nrow()

```

   
 
<br>
<br>
 
## Display N number for UTUVA data

<br>

```{r n_utuva}

tuva_utuva %>% filter(cohort == "1940-born UTUVA cohort") %>% nrow() 
 

```

   
 
<br>
<br>

 
## Display structure for TUVA data

<br>

```{r structure_tuva}

# Display data structure
tuva %>% str()

# Counts by educations groups
tuva %>% count(basic_school.factor3)
 

```

 
<br>
<br>


## ANOVA analyses for SBP within TUVA

<br>

```{r anova_tuva_sbp}


# SBP by basic school education
tuva %>% group_by(basic_school.factor3) %>%
         summarize(mean_SBP = mean(tuva1sbp)) %>%
         as.data.frame() %>% 
         mutate_if(is.numeric, round, 1)

# ANOVA
anova_sbp_tuva <- aov(tuva1sbp ~ basic_school.factor3, data = tuva)
summary_aov1 <- summary(anova_sbp_tuva)
print(summary_aov1)

```


<br>
<br>
 
## ANOVA analyses for DBP within TUVA

<br>

```{r anova_tuva_dbp}

# DBP by basic school education
tuva %>% group_by(basic_school.factor3) %>%
         summarize(mean_DBP = mean(tuva1dbp)) %>%
         as.data.frame() %>% 
         mutate_if(is.numeric, round, 1)

# ANOVA
anova_dbp_tuva <- aov(tuva1dbp ~ basic_school.factor3, data = tuva)
summary_aov2 <- summary(anova_dbp_tuva)

print(summary_aov2)

```


<br>
<br>
 


## Chi square test analyses for hypertension within TUVA

<br>

```{r tuva_ht}

# Summarize variation in antihypertensive medication use by basic school education
tuva %>%
    group_by(basic_school.factor3) %>%
    summarise(n_total = n(),
              n_ht = sum(hypertension == 1),
              percentage_ht = 100 * mean(hypertension))

tuva$hypertension.factor <- factor(tuva$hypertension)

# Chi-square test for association between antihypertensive medication use and education groups
chisq.test(table(tuva$hypertension.factor, tuva$basic_school.factor3))

```


<br>
<br>





## Chi square test analyses for BP meds within TUVA

<br>

```{r tuva_bp_meds}

# Summarize variation in antihypertensive medication use by basic school education
tuva %>%
    group_by(basic_school.factor3) %>%
    summarise(n_total = n(),
              n_on_BP_meds = sum(antihypertensive_med == 1),
              percentage_meds = 100 * mean(antihypertensive_med))


tuva$antihypertensive_med.factor <- factor(tuva$antihypertensive_med)

# Chi-square test for association between antihypertensive medication use and education groups
chisq.test(table(tuva$antihypertensive_med.factor, tuva$basic_school.factor3))

```


<br>
<br>





## Display structure for UTUVA data

<br>

```{r structure_utuva}

# Display data structure
utuva %>% str()

# Counts by educations groups
utuva %>% count(basic_school.factor3)


```

<br>
<br>


## ANOVA analyses for SBP within UTUVA

<br>

```{r anova_utuva_sbp}


# SBP by basic school education
utuva %>% group_by(basic_school.factor3) %>%
         summarize(mean_SBP = mean(utuva1sbp)) %>%
         as.data.frame() %>% 
         mutate_if(is.numeric, round, 1)

# ANOVA
anova_sbp_utuva <- aov(utuva1sbp ~ basic_school.factor3, data = utuva)
summary_aov3 <- summary(anova_sbp_utuva)
print(summary_aov3)

```


<br>
<br>


## ANOVA analyses for DBP within UTUVA

<br>

```{r anova_utuva_dbp}

# DBP by basic school education
utuva %>% group_by(basic_school.factor3) %>%
         summarize(mean_DBP = mean(utuva1dbp)) %>%
         as.data.frame() %>% 
         mutate_if(is.numeric, round, 1)

# ANOVA
anova_dbp_utuva <- aov(utuva1dbp ~ basic_school.factor3, data = utuva)
summary_aov4 <- summary(anova_dbp_utuva)
print(summary_aov4)

```


<br>
<br>


## Chi square test analyses for hypertension within UTUVA

<br>

```{r utuva_ht}

# Summarize variation in antihypertensive medication use by basic school education
utuva %>%
    group_by(basic_school.factor3) %>%
    summarise(n_total = n(),
              n_ht = sum(hypertension == 1),
              percentage_ht = 100 * mean(hypertension))


utuva$hypertension.factor <- factor(utuva$hypertension)

# Chi-square test for association between antihypertensive medication use and education groups
chisq.test(table(utuva$hypertension.factor, utuva$basic_school.factor3))

```


<br>
<br>



## Chi square test analyses for BP meds within UTUVA

<br>

```{r utuva_bp_meds}

# Summarize variation in antihypertensive medication use by basic school education
utuva %>%
    group_by(basic_school.factor3) %>%
    summarise(n_total = n(),
              n_on_BP_meds = sum(antihypertensive_med == 1),
              percentage_meds = 100 * mean(antihypertensive_med))


utuva$antihypertensive_med.factor <- factor(utuva$antihypertensive_med)

# Chi-square test for association between antihypertensive medication use and education groups
chisq.test(table(utuva$antihypertensive_med.factor, utuva$basic_school.factor3))

```


<br>
<br>



# TUVA mean SBP

<br>

```{r}

tuva %>%
  group_by(basic_school.factor3) %>%
  summarize(
    mean_tuva1sbp = format(round(mean(tuva1sbp, na.rm = TRUE), 1), nsmall = 1),
    sd_tuva1spd = sd(tuva1sbp, na.rm = TRUE)
  )
```

<br>
<br>

# TUVA mean DBP

<br>

```{r}

tuva %>%
  group_by(basic_school.factor3) %>%
  summarize(
    mean_tuva1dbp = mean(tuva1dbp, na.rm = TRUE),
    sd_tuva1dbp = sd(tuva1dbp, na.rm = TRUE)
  )
```

<br>
<br>

# UTUVA mean SBP

<br>

```{r}

utuva %>%
  group_by(basic_school.factor3) %>%
  summarize(
    mean_utuva1sbp = format(round(mean(utuva1sbp, na.rm = TRUE), 1), nsmall = 1),
    sd_utuva1sbp = sd(utuva1sbp, na.rm = TRUE)
    )
```


<br>
<br>

# UTUVA mean DBP

<br>

```{r}

utuva %>%
  group_by(basic_school.factor3) %>%
  summarize(
    mean_utuva1dbp = mean(utuva1dbp, na.rm = TRUE),
    sd_utuva1dbp = sd(utuva1dbp, na.rm = TRUE)
  )
```

<br>
<br>

# Summary of the ANOVA results

<br>

```{r}

anova_results <- data.frame(
  Model = c("Basic education TUVA: SBP", "Basic education TUVA: DBP", 
            "Basic education UTUVA: SBP", "Basic education TUVA: DBP"),
  F_statistic = c(summary_aov1[[1]]$`F value`[1],
                  summary_aov2[[1]]$`F value`[1],
                  summary_aov3[[1]]$`F value`[1],
                  summary_aov4[[1]]$`F value`[1]),
  p_value = c(summary_aov1[[1]]$`Pr(>F)`[1],
              summary_aov2[[1]]$`Pr(>F)`[1],
              summary_aov3[[1]]$`Pr(>F)`[1],
              summary_aov4[[1]]$`Pr(>F)`[1])
)


kable(anova_results, caption = "ANOVA Results Summary")


```

